sdmTMB: an R package for fast, flexible, and user-friendly generalized linear mixed effects models with spatial and spatiotemporal random fields

### 2022-11-02 doc/paper-jss/owls.Rmd
Edits
More caching, edit abstract

```{r print-fit-owls, warning=FALSE, echo=TRUE, message=FALSE, cache=FALSE}
print(fit_owl2)
```

@@

```{r tidy-owls, cache=FALSE}
tidy(fit_owl2, conf.int = TRUE)
tidy(fit_owl2, effects = "ran_pars", conf.int = TRUE)
```

@@

```{r theme-owls, cache=FALSE, echo=FALSE}
theme_set(theme_bw() + theme(
  plot.title = element_text(size = 12),
  legend.position = "top", legend.justification = c(0, 0), # move legend to top left
  legend.title = element_text(size = 10, hjust = 0), legend.key.height = unit(0.3, "cm"),
  strip.background = element_rect(fill = NA, colour = NA), # de-emphasize facet labels
  panel.grid.major = element_line(colour = "grey90"), # for lat/lons
  axis.title = element_blank() # lat/lon is clear from axis text
)) 
```

@@

```{r plot-map-ci, echo=FALSE, eval=TRUE, out.width="90%", cache=FALSE, fig.asp=0.3}
p1 <- ggplot(data = p_proj) +
  geom_point(
    data = p_mean,
    aes(X, Y, colour = nao_lwr, size = mean_est_count), alpha = 0.5
  ) +
  geom_sf(data = coast, colour = "gray50") +
  geom_sf(data = lakes, colour = "gray50", fill = NA) +
  coord_sf(
    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000), # adjusts space on sides
    ylim = c(min(p_proj$Y), max(p_proj$Y))
  ) +
  scale_colour_viridis_c(
    limit = c(min(p_mean$nao_lwr), max(p_mean$nao_upr)), 
    guide = guide_colourbar(direction = "horizontal", title.position = "top")
  ) +
  guides(size = "none", colour = "none") +
  # labs(colour = "Lower 90% CI") +
  ggtitle("Lower 80% CI") +
  theme(legend.position = c(0.1, 0.1), axis.title = element_blank())

@@

```{r plot-all-effects-owls, echo=FALSE, eval=TRUE, cache=FALSE, dpi=220}
ggplot(data = p_proj) +
  geom_point(data = p, aes(X, Y, colour = exp(est)), size = 0.4, alpha = 0.5) +
  coord_sf(
    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000),
    ylim = c(min(p_proj$Y), max(p_proj$Y))
  ) +
  scale_colour_viridis_c(
    trans = "sqrt", # makes variation in low numbers visible
    # can also trim the extreme high values
    limits = c(0, quantile(exp(p$est), 0.995)), na.value = "yellow",
    guide = guide_colourbar(title.position = "top")
  ) +
  labs(colour = "Predicted counts") +
  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
  facet_wrap(~year_f)
```

>>>>

```{r print-fit-owls, warning=FALSE, echo=TRUE, message=FALSE}
print(fit_owl2)
```

@@

```{r tidy-owls}
tidy(fit_owl2, conf.int = TRUE)
tidy(fit_owl2, effects = "ran_pars", conf.int = TRUE)
```

@@

```{r theme-owls, echo=FALSE}
theme_set(theme_bw() + theme(
  plot.title = element_text(size = 12),
  legend.position = "top", legend.justification = c(0, 0), # move legend to top left
  legend.title = element_text(size = 10, hjust = 0), legend.key.height = unit(0.3, "cm"),
  strip.background = element_rect(fill = NA, colour = NA), # de-emphasize facet labels
  panel.grid.major = element_line(colour = "grey90"), # for lat/lons
  axis.title = element_blank() # lat/lon is clear from axis text
)) 
```

@@

```{r plot-map-ci, echo=FALSE, eval=TRUE, out.width="90%",fig.asp=0.3}
p1 <- ggplot(data = p_proj) +
  geom_point(
    data = p_mean,
    aes(X, Y, colour = nao_lwr, size = mean_est_count), alpha = 0.5
  ) +
  geom_sf(data = coast, colour = "gray50") +
  geom_sf(data = lakes, colour = "gray50", fill = NA) +
  coord_sf(
    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000), # adjusts space on sides
    ylim = c(min(p_proj$Y), max(p_proj$Y))
  ) +
  scale_colour_viridis_c(
    limit = c(min(p_mean$nao_lwr), max(p_mean$nao_upr)), 
    guide = guide_colourbar(direction = "horizontal", title.position = "top")
  ) +
  guides(size = "none", colour = "none") +
  # labs(colour = "Lower 90% CI") +
  ggtitle("Lower 80% CI") +
  theme(legend.position = c(0.1, 0.1), axis.title = element_blank())

@@

```{r plot-all-effects-owls, echo=FALSE, eval=TRUE, dpi=220}
ggplot(data = p_proj) +
  geom_point(data = p, aes(X, Y, colour = exp(est)), size = 0.4, alpha = 0.5) +
  coord_sf(
    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000),
    ylim = c(min(p_proj$Y), max(p_proj$Y))
  ) +
  scale_colour_viridis_c(
    trans = "sqrt", # makes variation in low numbers visible
    # can also trim the extreme high values
    limits = c(0, quantile(exp(p$est), 0.995)), na.value = "yellow",
    guide = guide_colourbar(title.position = "top")
  ) +
  labs(colour = "Predicted counts") +
  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
  facet_wrap(~year_f)
```

### 2022-11-02 doc/paper-jss/paper-jss.Rmd
More caching, edit abstract

---
documentclass: jss
author:
  - name: Sean C. Anderson
    orcid: 0000-0001-9563-1937
    affiliation: 'Pacific Biological Station, Fisheries and Oceans Canada \AND'
    address: |
      | Pacific Biological Station, Nanaimo, BC, Canada
    email: \email{sean.anderson@dfo-mpo.gc.ca}
  - name: Eric J. Ward
    orcid: 0000-0002-4359-0296
    affiliation: 'Northwest Fisheries Science Center, National Marine Fisheries Service \AND'
  - name: Philina A. English
    orcid: 0000-0003-2992-6782
    address: ''
    affiliation: 'Pacific Biological Station, Fisheries and Oceans Canada \AND'
  - name: Lewis A. K. Barnett
    orcid: 0000-0002-9381-8375
    affiliation: 'Alaska Fisheries Science Center, National Marine Fisheries Service'
bibliography: refs.bib
title:
  formatted: "\\pkg{sdmTMB}: An \\proglang{R} Package for Fast, Flexible, and User-Friendly Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields"
  # If you use tex in the formatted title, also supply version without
  plain:     "sdmTMB: An R Package for Fast, Flexible, and Accessible Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields"
  # For running headers, if needed
  short:     "\\pkg{sdmTMB}: geostatistical SPDE-based GLMMs with \\pkg{TMB}"
abstract: >
  Geostatistical data are pervasive and available in increasingly large volumes. However, appropriate models to analyse these data, such as generalised linear mixed effects models (GLMMs) with Gaussian random fields, are computationally intensive and challenging to implement. Here, we introduce the \proglang{R} package \pkg{sdmTMB}, which extends the flexible interface familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} to include spatial and spatiotemporal latent fields using a predictive-process SPDE-(stochastic partial differential equation) based approach. SPDE matrices are constructed with \pkg{INLA} and estimation is conducted via maximum marginal likelihood with Template Model Builder (\pkg{TMB}), although penalised likelihood and Bayesian inference (via \pkg{tmbstan} and \pkg{rstan}) are possible. We describe the statistical model, illustrate the package through two case studies, and compare the functionality, speed, and interface to related software. The case studies illustrate \pkg{sdmTMB}'s flexibility and useability by implementing penalised smoothers, non-stationary processes (time-varying and spatially varying coefficients), and anisotropy (asymmetry in spatial correlations). Additional functionality includes delta (hurdle) models, break-point effects, and out-of-sample cross validation. Finally, we demonstrate that \pkg{sdmTMB} is 4--16 times faster than \pkg{INLA} or \pkg{inlabru}. \pkg{sdmTMB}'s accessible interface has already opened this useful class of models to a wider audience within ecological species distribution modelling (SDMs). We anticipate its application to the full spectrum of geostatistical data will prove equally fruitful.
keywords:
  # at least one keyword must be supplied
  formatted: [test]
  plain:     [test]
preamble: >
  \usepackage{amsmath}
  \usepackage{amssymb}
  \usepackage{bm}
  \usepackage{lscape}
  \usepackage{tablefootnote}
  \usepackage{threeparttable}
  \usepackage{booktabs}
output: rticles::jss_article
---

>>>>

---
documentclass: jss
author:
  - name: Sean C. Anderson
    orcid: 0000-0001-9563-1937
    affiliation: 'Pacific Biological Station, Fisheries and Oceans Canada \AND'
    address: |
      | Pacific Biological Station, Nanaimo, BC, Canada
    email: \email{sean.anderson@dfo-mpo.gc.ca}
bibliography: refs.bib
title:
  formatted: "\\pkg{sdmTMB}: An \\proglang{R} Package for Fast, Flexible, and User-Friendly Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields"
  # If you use tex in the formatted title, also supply version without
  plain:     "sdmTMB: An R Package for Fast, Flexible, and Accessible Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields"
  # For running headers, if needed
  short:     "\\pkg{sdmTMB}: geostatistical SPDE-based GLMMs with \\pkg{TMB}"
abstract: >
  Geostatistical data are pervasive and available in increasingly large volumes. However, appropriate models to analyse these data, such as generalised linear mixed effects models (GLMMs) with Gaussian random fields, are computationally intensive and challenging to implement. Here, we introduce the \proglang{R} package \pkg{sdmTMB}, which extends the flexible interface familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} to include spatial and spatiotemporal latent fields using a predictive-process SPDE-(stochastic partial differential equation) based approach. SPDE matrices are constructed with \pkg{INLA} and estimation is conducted via maximum marginal likelihood with Template Model Builder (\pkg{TMB}) or Bayesian inference via \pkg{tmbstan} and \pkg{rstan}). We describe the model and use case studies to illustrate sdmTMB's flexibility by implementing penalised smoothers, non-stationary processes (time-varying and spatially varying coefficients), and anisotropy (asymmetry in spatial correlations). Additional functionality includes delta (hurdle) models, break-point effects, and out-of-sample cross validation. Finally, we compare the functionality, speed, and interfaces of related software, demonstrating that sdmTMB is 4–16 times faster than INLA. We anticipate sdmTMB's accessible interface will help open this useful class of models to the full spectrum geostatistical analysts.
keywords:
  # at least one keyword must be supplied
  formatted: [test]
  plain:     [test]
preamble: >
  \usepackage{amsmath}
  \usepackage{amssymb}
  \usepackage{bm}
  \usepackage{lscape}
  \usepackage{tablefootnote}
  \usepackage{threeparttable}
  \usepackage{booktabs}
output: rticles::jss_article
---
